Measuring the double-drift illusion with hand tracking: capturing illusion resets

### 2020-10-05 Manuscript.Rmd
edits to writing

---
title: "Measuring the double-drift illusion with hand tracking: capturing illusion resets"
author: 
- Bernard Marius 't Hart 1,*
- Denise Y.P. Henriques 1
- Patrick Cavanagh 1,2,3
output:
  pdf_document: default
  word_document: default
  html_document:
    df_print: paged
header-includes:
- \usepackage{lineno}
- \linenumbers
- \usepackage{caption}
- \captionsetup[figure]{labelformat=empty}
bibliography: bibliography.bib
csl: journal-of-vision.csl
link-citations: yes
---

@@

In this experiment, two tasks were performed by two separate groups of participants. Both tasks asked participants to use a stylus on a drawing tablet to track a moving gabor with some amount of internal motion that elicited the double-drift illusion. We test if tracking the illusion continuously, prevents or decreases the illusion.

In the first task, participants were asked to continuously track the position of a gabor that moved forward and backward along the Y-axis (of the horizontal screen) for the duration of 12 _s_, using a stylus on a drawing tablet. Before trial onset, the participant moved the stylus to the middle of the screen ( where the gabor would appear ), and was subsequently asked to fixate a point on the left or right of the screen, and given 1.5 _s_ to do so. Since the screen was seen via a mirror, the stimulus appeared to move forward and backward on a horizontal plane at the same depth as the drawing tablet. The internal motion of the gabor was inverted at the far and close ends of the physical path, where external motion was also inverted. Internal motion could be 3, 1, 0, -1 or -3 _cps_. Since each moving gabor was shown for 12 _s_, and it moved from one end to another in 2 _s_, it completed 5 full and 2 half passes of the path.

In the second task, only the highest internal motion was used (±3 _cps_), trials include only 3 complete passes, but these took 4 _s_ each. As before, the fixation point (left or right) was random for each trial, while initial external motion (forward or backward) as well as internal motion were counterbalanced across trials.

@@

In a second task, we only used the higher internal speed (±3 _cps_) and increased the duration of a single pass of the stimulus from 2 _s_ to 4 _s_, both to increase the occurence of spontaneous resets. Average headings and heading distributions for 6 participants can be seen in Figure 3. Despite lower external motion speed, the illusion is as strong as in the conditions with the same internal motion in the first task.

```{r fig3, echo=FALSE, message=FALSE, fig.width=8, fig.height=4, fig.cap="\\label{fig:fig3}**Figure 3. Heading in Experiment 1b.** Average heading and heading distribution across the middle 3 _s_ of each pass. Distance from origin is time, angle is angle. In dark blue lines the average heading, in light blue a polar heatmap giving an impression of the distribution of heading. **Left:** Data for all participants. **Right:** Group average."}
plotBoundedTracking_V2()
```

### Illusion strength

The illusion is clearly there in both tasks, but does tracking decrease the strength of the illusion compared to simply observing it? We compare illusion strength to that reported previously [@Cavanagh2019] in Figure 4. As can be seen the averages from our data here (in red) coincide almost exactly with the illusion strength predicted by the model. That is; tracking the illusion while observing it does not change the strength of the illusion.

```{r fig4, echo=FALSE, message=FALSE, fig.width=8, fig.height=5, fig.cap="\\label{fig:fig4}**Figure 4. Illusion strength.** The strength of the illusion in degrees angle deviation from the physical path is plotted over external motion speed ( _dva/s_ ). In red are the average illusions strengths for ±3 _cps_, ±1 _cps and 0 cps. The black circles and line are the data and model from Cavanagh & Tse (2019). The grey lines are the illusion strengths their model predicts for the internal motion speeds used here."}
plotStrengthComparison()
```

### Resets

In order to detect potential resets in the first task, we consider that in the 0 internal motion condition there is no illusion, so that we measure the tracking of the veridical path of the stimulus. Since both strong, "jump" resets and weak "hit-the-wall" resets (see Fig 1, bottom left) would change percieved direction back to or crossing the veridical motion path, their smooth-spline interpolated paths would contain samples similar in heading to the 0 internal motion condition. We use a 95% confidence interval of heading in the 0 internal motion condition to select trajectories from conditions with non-zero heading that may show a reset. These trajectories are drawn in red (only 1 per condition/participant), but as can be seen, the deviations from the condition's dominant heading occur just before or after the external motion reversals, so they may simply reflect difficulty of tracking the stimulus at the reversal points, rather than resets.

The second task does not give any additional information.

Except for the first part of participant 12's trajectories, all heading data matched that of the 3 _cps_ condition in experiment 1a. Those first few samples in participant 12 could reflect a participant specific delay in tracking the stimulus position. In other words, no unambiguous resets were captured in either experiment. It is possible that resets were perceived but were too quick to respond to in the continuous tracking task. Alternatively, tracking may have stabilized the illusion, preventing resets from occurring. Hence, we performed a second experiment that would counter both explanations for the absence of (spontaneous) resets.

@@

In a second experiment, we ask participants to re-trace the perceived path of a double-drift stimulus, immediately after viewing the stimulus. This allows participants more time to carefully reproduce the path. We use the recorded trajectories to assess points where resets occured. The reset points can then be used to determine if resets are time-limited or space-limited.

@@

Here we used a re-tracing procedure where participants drew the perceived path of a double-drift stimulus making a single pass across the workspace. This provided more time for the participants to reproduce any resets, i.e. without the need for real-time tracking. Nakayama & Holcombe [-@Nakayama2020] have also used a re-tracing paradigm, in their case to study triggered resets. The re-tracing task was done in half of 8 blocks. In the other half, participants oriented a line to indicate the initial perceived direction of the stimulus motion. Block types were alternated, and the order was counterbalanced across participants. Each block used the 6 combinations of two external speeds ( corresponding to 3 or 4 _s_ presentation time ), and three internal speeds ( 2, 3 or 4 _cps_ ), 6 times. Within each of the two tasks, each of the combinations of internal and external speed was presented 24 times in total.

In both kinds of trials, participants first had to move the stylus to the start position of the gabor, and then fixate a point to the left or right of the centre of the workspace. The double-drift gabor started at the near end of the display, and moved along the horizontal display away from the participant, always making a single, forward pass. After, the gabor had disappeared, participants could respond in one of two ways. When re-tracing the perceived path with the stylus on the tablet, the drawn path would show up as a red line. Participants could "key in" their response by pressing ENTER, or start over by pressing ESCAPE. When indicating only the initial direction of movement, participants would orient a line ( 75 _px_ or ~3 _cm_ ) originating from the starting point of the gabor’s path. They could change the orientation by pressing either the LEFT ARROW key to orient the line more clockwise or the RIGHT ARROW key to orient the line more counter clockwise. They could "key in" their response by pressing ENTER. In both tasks, there was no time-limit for the response, and careful responding was encouraged.

@@

In Figure 5, for each of the 6 combinations of internal speed (2, 3 or 4 _cps_) and external speed (3 or 4 _s_ stimulus pass) we show the data for one example participant, as well as the heading distribution for all participants. While the heading distributions don't show resets, for many of the participants, we can see that the re-traced paths do show resets, many of which are detected by our algorithm.

```{r fig5, echo=FALSE, message=FALSE, fig.width=8, fig.height=5, fig.cap="\\label{fig:fig5}**Figure 5. Example Trajectories and Heading in Experiment 2.** For each of the six conditions, defined by 3 internal speeds (2, 3 and 4 _cps_) and two external speeds (corresponding to 3 and 4 _s_ passes), we show one example participant's trajectories and reset points, as well as the heading distribution."}
plotExampleData()
```

On average, the algorithm detects resets in 75% of trials. For three participants, only about half the trials show resets (42%, 44% and 58%). For the remaining six participants the algorithm detects resets in 72%-99% of trials, with an average of 88%. The minimum number of resets in one of the six conditions for any participant is 7 (out of 24), with the average at 17. Since participants experiencing a "hit-the-wall reset" could display a direction change in their paths only by chance, 50% of trials with resets seems reasonable. Neither percept is incorrect, and the average reset positions still indicates some limit to the illusion. Therefore, we keep all participants and resets.

First, we consider the line angle responses as a measure of illusion strength. The strength of the illusion should depend on both the internal and external speed of the double drift stimulus. As can be seen in Figure 6 (left), at least the internal speed, and perhaps the external speed may predict illusion strength as measured by line angles. To test this, we perform a mixed-design ANOVA on line angle, using both internal and external speed as within-subject factors. We find main effects of both internal speed ( _F_(2,16)=12.7; _p_<.001, $\eta^2$=0.049) and external speed ( _F_(1,8)=6.89; _p_=.03; $\eta^2$=0.031), and no interaction. That is: we verify that illusion strength varied systematically with our manipulations of the stimulus.

```{r fig6, echo=FALSE, message=FALSE, fig.width=8, fig.height=4, fig.cap="\\label{fig:fig6}**Figure 6. Measures of Illusion Strength** Line angle responses and reset point direction should both measure illusion strength. **Left:** Line Angle response as a function of internal and external speed, averaged across 9 participants in each of the 6 conditions. Shaded areas denote 95% confidence intervals. **Right:** Angles of samples halfway to a reset over line angle response. Each dot corresponds to one participants average in one of the 6 conditions. Diagonal line indicates a weighted regression (weighted by one over the square of the line angle response to counter heteroscedasticity), and the shaded area shows a 95% confidence interval for the regression."}
plotValidity()
```

We also take the angle of the sample on each re-tracing trajectory that is halfway between the origin and the detected reset, as an alternative measure of illusion strength. If both the angle half-way to reset and line angle responses are good measures of illusion strength, they should be related. As can be seen from the data in Figure 6 (right), there might be a relation between the two variables, but there is also heteroscadasticity. We counter this by doing a weighted regression, with the weights set to one over the squared line angle ($\frac{1}{\alpha^2}$). Reported line angles predicted reset point direction (slope=0.71; _t_(49)=4.54, _p_<.001; $R^2$=.282; _F_(1,49)=20.6), verifying that the two measures - while imperfect - are likely both indicators of double-drift illusion strength. From here on, we will only use the angle of samples half-way to detected reset points as measure of illusion strength.

@@

Now we look at the distribution of reset points, and whether they can be explained by a temporal or spatial limit. In Figure 7 (left), we can see that the reset points vary more along the Y-axis than the X-axis. That is, predicting reset points from a single Y coordinate (time-cutoff) would leave larger residuals than predicting reset points from a single X coordinate (or spatial cutoff). In fact, the standard deviation of reset point X-coordinates is about 21.9% of teh standard deviation of reset point Y-coordinates. This already indicates that a spatial limit explains about 4 times more of the variance in reset points than a temporal limit.

```{r fig7, echo=FALSE, message=FALSE, fig.width=8, fig.height=6, fig.cap="\\label{fig:fig7}**Figure 7. Reset Points and Single Limits** **Left:** The average reset points for each participant and stimulus condition displayed spatially. Y-coordinates have been converted to time, in order to normalize for the two external speeds. Transparent purple lines indicate the standard deviation of coordinates. Dashed vertical red line indicates the median X-coordinate (1.6 _cm_). Dashed horizontal blue line indicates the median Y-coordinate (1.4 _s_). **Top right:** X-coordinates predicted from a single temporal limit. **Bottom right:** Y-coordinates predicted from a single spatial limit."}
plotResetPoints()
```

Nevertheless, we test two simple models, based on straighforward trigonometry. First, we calculate the slope of a line through the origin and a sample halfway to the reset point. Then, if the perceived path of the gabor falls on a straight line, and reset points occur at a set limit, then a temporal limit would predict X coordinates:

@@

And a spatial limit would predict Y coordinates:

@@

Where $L_y$ and $L_x$ are a temporal (y) and spatial (x) limit that are free parameters, and s corresponds to the slope. If we fit these simple models by minimizing the mean square error (as the distance in centimeters between predicted and actual reset points), the optimal parameters indicate a temporal limit of ~1.4 _s_ (AIC=258.4) and a spatial limit of ~1.7 _cm_ (AIC=217.7). This is close to the median values of the respective coordinates. The model with the spatial limit fits the data better than the model with a temporal limit ( _p_<.001, relative log-likelihood based on AICs). This again indicates that a spatial limit explains more of the data.

However, it may be that after crossing a spatial limit, it takes some given amount of time before a reset of the illusion is perceived. This would mean that both limits explain separate parts of the data. We model this by finding optimal parameter values for both a spatial limit and a temporal limit simultaneously, when minimizing the distance between predicted and atcual reset points (in centimeters). The equations in the model are:

@@

Figure 8 shows both the actual reset point coordinates (dots), given the illusion strength as the angle to a sample half-way to the reset point, as well as the model's predictions of reset point coordinates (lines).

```{r fig8, echo=FALSE, message=FALSE, fig.width=6, fig.height=6, fig.cap="\\label{fig:fig8}**Figure 8. Reset Points and Combined Limits**"}
plotSequentialModel()
```

The limits this model generates are a 1.42 _cm_ spatial limit and a 0.28 _s_ temporal limit. While the MSE and AIC of the model are lower than that of the model with the single spatial limit, it does not explain the data better ( _p_=.53). This means that a simple spatial limit accounts for most of the data. It is possible that the way we've modeled the additional temporal delay after crossing the spatial limit is not accurate, as it may follow a poisson distribution, which we do not account for here. However, regardless of whether or not a temporal limit adds anything, a spatial limit appears to explain the majority of spontaneous double-drift resets.

```{r comment, echo=FALSE, message=FALSE}
#It may be that during online tracking, proprioceptive information on the location of the hand is included in the estimate of the location of the stimulus. This may have decreased the likelihood of a reset, and may also changed the strength of the illusion. Since some participants did both experiments (X/9), we can at least test if illusion strength is different for online tracking and offline re-tracing.
```

@@

While here we find that a spatial limit is the better explanation for resets, other evidence may point at a temporal limit instead. In an as of yet unpublished study [Sirui Liu] obtained the point of subjective equivalence (PSE) of a double drift stimulus, and a static point that appeared some time after the double drift stimulus disappeared. The double-drift stimulus was on a 2 or 4 DVA length trajectory, traversed in 1, 2 or 3 seconds. The PSE's can be modelled very accurately with a model that assumes a fixed chance of a reset per unit of time (~0.7%/s), that is: it explains the data using only a temporal bound on the illusion. Interestingly, this fully contradicts the current findings. The model predicts that the PSE decreases over time as it becomes more likely that a reset has occurred. We will compare the two data sets here (see Fig 9).

```{r fig9, echo=FALSE, message=FALSE, fig.width=6, fig.height=8, fig.cap="\\label{fig:fig9}**Figure 9. PSE's of double-drift position and continuous tracking**"}
plotAverageTrajectories()
```

If the PSE's express the average perceived deviations from the actual trajectory of the double-drift stimulus, then they should show a similar pattern as the averaged trajectories in experiment 2. In order to be able to compare the two data sets, we will include all trajectories, including those where our simple algorithm did not detect a reset, as such trials should also be included in the PSE data. Further, we scale the trajectories on both axes, so that the Y end-point matches the duration of stimulus. Since the PSE data was collected in degrees visual angle, we convert the unit to degrees visual angles, using the estimate at the centre of the workspace. For the 4-second stimuli, we removed the last second. Figure 9 shows the individual participants' and the averaged trajectories (purple, red, orange) as well as the PSE's obtained in the other study (shown separately for 2 and 4 DVA stimulus trajectories) in gray.

For all internal motion speeds, we see a very similar average trajectory, that does not get closer to veridical over time. We verified this with a repeated measures ANOVA on perceived deviation from veridical stimulus position, using internal motion speed (2, 3, or 4 cycles/s) and time (1, 2, or 3 seconds after stimulus onset) as factors. If the trajectory data from experiment 2 here follows the same pattern as the PSE data, there should be an effect of time, and perhaps an interaction with internal motion speed. We find no effect of time ( _F_(2,16)=0.011, _p_=.957, $\eta^2$<.001), no effect of internal motion speed ( _F_(2,16)=0.127, _p_=.814, $\eta^2$=.0015) and no interaction ( _F_(4,32)=0.361, _p_=.635, $\eta^2$=.002). This means that we can't replicate the pattern found with the PSE approach. There may be several reasons for this. First, we notice that the trajectory data is highly dependent on participants. It could be that the re-tracing method provides very noisy estimates, hiding any effect. It could also be that the effects truly vary a lot between participants. Since we can see that some participants do seem to follow the pattern found with the PSE approach, and some show a pattern of increasing average deviation between perceived and true position, it may that the differences between the data sets can simply be explained by random sampling of relatively small numbers of participants.

@@

We first tested if tracking the position of a double-drift stimulus changes the illusory percept of it's position, and found that it doesn't. We then set out to test if spontaneous resets in the double-drift illusion reflect discrepancies between two position signals, occurring at some given size of this discrepancy, or if they simply occur after some time elapses, analogous to other illusions or bi-stable stimuli. To measure the location of resets we first ask participants to track the perceived location of the double-drift stimulus online. While the heading of the tracking movements is sensitive to the strength of the illusion, there are no clear reset points in the tracking data. In a second experiment we ask participants to draw the perceived path of the stimulus after a single pass. In this re-tracing experiment the drawn traces reflect the strength of the illusion just as well, and in most trials there is a clear reset point, for either a jump reset or a hit-the-wall reset. The positions of these reset points do no vary much with elapsed time, but are better explained by the distance from the true position. This has implications for the way position perception relies on motion signals.

@@

Some of our participants seem to perceive "jump resets" though, and the model as published, does not predict any such resets. This could be added by including a causal inference step [e.g. @Wei2011]. For the subset of participants that experience jump resets, there may be a process that compares the two position signals (perceived and retinal) and when these exceed some threshold, decides that the two position signals do not share a common source. At that point the perceived position may be discarded, or reset to the retinal position, and the build-up of the illusion may start again. For the remaining participants, either the discrepancy between retinal and perceived location never exceeds their threshold, or there may be another reason they do not employ any causal inference process.

Recently, it has been suggested that resets can be triggered by planned eye-movements to the stimulus [@Nakayama2020], particularly "jump" resets. In this experiment, we did not control for eye-movements with an eye-tracker, so we can't directly investigate this idea. Given a pre-motor theory of attention like mechanism, when spontaneously planned eye-movements trigger (jump) resets, such resets could occur at any point during the stimulus pass. However, it seems that for the participants that display "jump" resets in their re-tracing data, these happen at a fairly consistent location, which suggests that jump resets are instead stimulus triggered. This should however be tested with eye-tracking data, so that this will remain a speculation here.

>>>>

---
title: "Measuring the double-drift illusion with hand tracking: capturing illusion resets"
author: 
- Bernard Marius 't Hart 1,*
- Denise Y.P. Henriques 1
- Patrick Cavanagh 1,2,3
output:
  word_document: default
  html_document:
    df_print: paged
  pdf_document: default
header-includes:
- \usepackage{lineno}
- \linenumbers
- \usepackage{caption}
- \captionsetup[figure]{labelformat=empty}
bibliography: bibliography.bib
csl: journal-of-vision.csl
link-citations: yes
---

@@

In this experiment, participants were asked to use a stylus on a drawing tablet to track a moving gabor with some amount of internal motion that elicited the double-drift illusion. We test if tracking the illusion continuously is a good measure of illusion strength, or if it prevents or decreases the illusion.

## Task

In this task, participants were asked to continuously track the position of a gabor that moved forward and backward along the Y-axis (of the horizontal screen) for the duration of 12 _s_, using a stylus on a drawing tablet. Before trial onset, the participant moved the stylus to the middle of the screen ( where the gabor would appear ), and was subsequently asked to fixate a point on the left or right of the screen, and given 1.5 _s_ to do so. Since the screen was seen via a mirror, the stimulus appeared to move forward and backward on a horizontal plane at the same depth as the drawing tablet. The internal motion of the gabor was inverted at the far and close ends of the physical path, where external motion was also inverted. Internal motion could be 3, 1, 0, -1 or -3 _cps_. Since each moving gabor was shown for 12 _s_, and it moved from one end to another in 2 _s_, it completed 5 full and 2 half passes of the path.

@@

## Illusion strength

The data clearly show the illusion, but does tracking decrease the strength of the illusion compared to simply observing it? We compare illusion strength to that reported previously [@Cavanagh2019] in Figure 3. As can be seen the averages from our data here (in red) coincide almost exactly with the illusion strength predicted by the model. That is; tracking the illusion while observing it does not change the strength of the illusion.

```{r fig3, echo=FALSE, message=FALSE, fig.width=8, fig.height=5, fig.cap="\\label{fig:fig3}**Figure 3. Illusion strength.** The strength of the illusion in degrees angle deviation from the physical path is plotted over external motion speed ( _dva/s_ ). In red are the average illusions strengths for ±3 _cps_, ±1 _cps and 0 cps. The black circles and line are the data and model from Cavanagh & Tse (2019). The grey lines are the illusion strengths their model predicts for the internal motion speeds used here."}
plotStrengthComparison()
```

## Resets

In order to detect potential resets, we consider that in the 0 internal motion condition there is no illusion, so that we measure the tracking of the veridical path of the stimulus. Since both strong, "jump" resets and weak "hit-the-wall" resets (see Fig 1, bottom left) would change percieved direction back to or crossing the veridical motion path, their smooth-spline interpolated paths would contain samples similar in heading to the 0 internal motion condition. We use a 95% confidence interval of heading in the 0 internal motion condition to select trajectories from conditions with non-zero heading that may show a reset. These trajectories are drawn in red in Figure 2 (for clarity only 1 per condition/participant is highlighted), but as can be seen, the deviations from the condition's dominant heading occur just before or after the external motion reversals, so they may simply reflect difficulty of tracking the stimulus at the reversal points, rather than resets.

No unambiguous resets were captured in either experiment. It is possible that resets were perceived but were too quick to respond to in the continuous tracking task. Alternatively, tracking may have stabilized the illusion, preventing resets from occurring. Either way, we performed a second experiment that would counter both explanations for the absence of (spontaneous) resets.

@@

In a second experiment, we ask participants to re-trace the perceived path of a double-drift stimulus, immediately after viewing the stimulus, as used by Nakayama & Holcombe [-@Nakayama2020]. This allows participants more time to carefully reproduce the path. We use the recorded trajectories to assess points where resets occured. The reset points can then be used to determine if resets are time-limited or space-limited.

@@

Here we used a re-tracing procedure where participants drew the perceived path of a double-drift stimulus making a single pass across the workspace. This provided more time for the participants to reproduce any resets, i.e. without the need for real-time tracking. The re-tracing task was done in half of 8 blocks. In the other half, participants oriented a line to indicate the initial perceived direction of the stimulus motion. Block types were alternated, and the order was counterbalanced across participants. Each block used the 6 combinations of two external speeds ( corresponding to 3 or 4 _s_ presentation time ), and three internal speeds ( 2, 3 or 4 _cps_ ), 6 times. Within each of the two tasks, each of the combinations of internal and external speed was presented 24 times in total.

In both kinds of trials, participants first had to move the stylus to the start position of the gabor, and then fixate a point to the left or right of the centre of the workspace. The double-drift gabor started at the near end of the display, and moved along the horizontal display away from the participant, always making a single, forward pass. After the gabor had disappeared, participants could respond in one of two ways. When re-tracing the perceived path with the stylus on the tablet, the drawn path would show up as a red line. Participants could "key in" their response by pressing ENTER, or start over by pressing ESCAPE. When indicating only the initial direction of movement, participants would orient a line ( 75 _px_ or ~3 _cm_ ) originating from the starting point of the gabor’s path. They could change the orientation by pressing either the LEFT ARROW key to orient the line more clockwise or the RIGHT ARROW key to orient the line more counter clockwise. They could "key in" their response by pressing ENTER. In both tasks, there was no time-limit for the response.

@@

In Figure 4, for each of the 6 combinations of internal speed (2, 3 or 4 _cps_) and external speed (3 or 4 _s_ stimulus pass) we show the data for one example participant, as well as the heading distribution for all participants. While the heading distributions don't show resets, for many of the participants, we can see that the re-traced paths do show resets, many of which are detected by our algorithm.

```{r fig4, echo=FALSE, message=FALSE, fig.width=8, fig.height=5, fig.cap="\\label{fig:fig4}**Figure 4. Example Trajectories and Heading in Experiment 2.** For each of the six conditions, defined by 3 internal speeds (2, 3 and 4 _cps_) and two external speeds (corresponding to 3 and 4 _s_ passes), we show one example participant's trajectories and reset points, as well as the heading distribution."}
plotExampleData()
```

On average, the algorithm detects resets in ~75% of trials. For three participants, only about half the trials show resets (42%, 44% and 58%). For the remaining six participants the algorithm detects resets in 72%-99% of trials, with an average of 88%. The minimum number of resets in each of the six conditions for any participant is 7 (out of 24 trials), with the average at ~17. Since participants experiencing a "hit-the-wall reset" (as opposed to a "jump" reset) might not always display a direction change in their paths, 50% of trials with resets seems reasonable. Either type of perceived resets appears valid, and the average reset positions still indicates some limit to the illusion. Therefore, we keep all participants and resets.

First, we consider the line angle responses as a measure of illusion strength. The strength of the illusion should depend on both the internal and external speed of the double drift stimulus. As can be seen in Figure 5 (left), at least the internal speed, and perhaps the external speed may predict illusion strength as measured by line angles. To test this, we perform a mixed-design ANOVA on line angle, using both internal and external speed as within-subject factors. We find main effects of both internal speed ( _F_(2,16)=12.7; _p_<.001, $\eta^2$=0.049) and external speed ( _F_(1,8)=6.89; _p_=.03; $\eta^2$=0.031), and no interaction. That is: we verify that illusion strength varied systematically with our manipulations of the stimulus.

```{r fig5, echo=FALSE, message=FALSE, fig.width=8, fig.height=4, fig.cap="\\label{fig:fig5}**Figure 5. Measures of Illusion Strength** Line angle responses and reset point direction should both measure illusion strength. **Left:** Line Angle response as a function of internal and external speed, averaged across 9 participants in each of the 6 conditions. Shaded areas denote 95% confidence intervals. **Right:** Angles of samples halfway to a reset over line angle response. Each dot corresponds to one participants average in one of the 6 conditions. Diagonal line indicates a weighted regression (weighted by one over the square of the line angle response to counter heteroscedasticity), and the shaded area shows a 95% confidence interval for the regression."}
plotValidity()
```

As specified above, we also take the angle of the sample on each re-tracing trajectory that is halfway between the origin and the detected reset, as an alternative measure of illusion strength. If both the angle half-way to reset and line angle responses are good measures of illusion strength, they should be related. As can be seen from the data in Figure 5 (right), there might be a relation between the two variables, but there is also heteroscadasticity. We counter this by doing a weighted regression, with the weights set to one over the squared line angle ($\frac{1}{\alpha^2}$). Reported line angles predicted reset point direction (slope=0.71; _t_(49)=4.54, _p_<.001; $R^2$=.282; _F_(1,49)=20.6), verifying that the two measures - while imperfect - are likely both indicators of double-drift illusion strength. From here on, we will only use the angle of samples half-way to detected reset points as measure of illusion strength.

@@

Now we look at the distribution of reset points, and whether they can be explained by a temporal or spatial limit. In Figure 7 (left), we can see that the reset points vary more along the Y-axis than the X-axis. That is, predicting reset points from a single Y coordinate (time-cutoff) would leave larger residuals than predicting reset points from a single X coordinate (or spatial cutoff). In fact, the standard deviation of reset point X-coordinates is about 21.9% of the standard deviation of reset point Y-coordinates. This already indicates that a spatial limit explains about 4 times more of the variance in reset points than a temporal limit.

```{r fig6, echo=FALSE, message=FALSE, fig.width=8, fig.height=6, fig.cap="\\label{fig:fig6}**Figure 6. Reset Points and Single Limits** **Left:** True path of the doubel-drift stimulus as gray arrow. The average reset points for each participant and stimulus condition displayed spatially, with the internal motion normalized to go rightward. Y-coordinates have been converted to time ( _s_ ), in order to normalize for the two external speeds. Transparent purple lines indicate the standard deviation of coordinates. Dashed vertical red line indicates the median X-coordinate (1.6 _cm_). Dashed horizontal blue line indicates the median Y-coordinate (1.4 _s_). **Top right:** X-coordinates predicted from a single temporal limit. **Bottom right:** Y-coordinates predicted from a single spatial limit."}
plotResetPoints()
```

Nevertheless, we test two simple models, based on straighforward trigonometry. First, we calculate the slope of a line ($s$) through the origin and a sample halfway to the reset point. This estimates illsuion strength as angular deviation from the true, straight path. If the perceived path of the gabor is approximated by a straight line (up to a reset), and reset points occur at a set limit, then a temporal limit ($L_y$) would predict X coordinates:

@@

And a spatial limit ($L_x$) would predict Y coordinates:

@@

Where $L_y$ and $L_x$ are a temporal (y) and spatial (x) limit that are free parameters, and s corresponds to the calculated slope. If we fit these simple models by minimizing the mean square error (as the absolute distance in centimeters between predicted and actual reset points), the optimal parameters indicate a temporal limit of ~1.4 _s_ (AIC=258.4) and a spatial limit of ~1.7 _cm_ (AIC=217.7). This is close to the median values of the respective coordinates. The model with the spatial limit fits the data better than the model with a temporal limit ( _p_<.001, relative log-likelihood based on AICs). This again indicates that a spatial limit explains more of the data.

However, it may be that after crossing a spatial limit, it takes some given amount of time before a reset of the illusion is perceived (or reported). This would mean that both limits explain separate parts of the data. We model this by finding optimal parameter values for both a spatial limit and a temporal limit simultaneously, when minimizing the distance (in _cm_ ) between predicted and actual reset points. The equations in the model are:

@@

Figure 7 shows both the actual reset point coordinates (dots), given the illusion strength as the angle to a sample half-way to the reset point, as well as the model's predictions of reset point coordinates (lines).

```{r fig7, echo=FALSE, message=FALSE, fig.width=6, fig.height=6, fig.cap="\\label{fig:fig7}**Figure 7. Reset Points and Combined Limits**"}
plotSequentialModel()
```

The limits that optimally fit the model to the data are a 1.42 _cm_ spatial limit and a 0.28 _s_ temporal limit. While the MSE and AIC of the model are lower than that of the model with the single spatial limit, it does not explain the data better ( _p_=.53). This means that a simple spatial limit accounts for most of the data. It is possible that the way we've modeled the additional temporal delay after crossing the spatial limit is not accurate, as it may not follow a normal distribution, which we do not account for here. However, regardless of whether or not a temporal limit adds anything, a spatial limit appears to explain the majority of spontaneous double-drift resets.

@@

While here we find that a spatial limit is the better explanation for resets, other evidence may point at a temporal limit instead. In an as of yet unpublished study [Sirui Liu] obtained the point of subjective equivalence (PSE) of a double-drift stimulus, and a static point that appeared some time after the double drift stimulus disappeared. The double-drift stimulus was on a 2 or 4 _dva_ length trajectory, traversed in 1, 2 or 3 _s_. The PSE's can be modelled very accurately with a model that assumes a fixed chance of a reset per unit of time (~0.7% / _s_), that is: it explains the data using only a temporal bound on the illusion. Interestingly, this fully contradicts the current findings. The model predicts that the PSE decreases over time as it becomes more likely that a reset has occurred. We will compare the two data sets here (see Fig 8).

```{r fig8, echo=FALSE, message=FALSE, fig.width=6, fig.height=8, fig.cap="\\label{fig:fig8}**Figure 8. PSE's of double-drift position and continuous tracking**"}
plotAverageTrajectories()
```

If the PSE's express the average perceived deviations from the actual trajectory of the double-drift stimulus, then they should show a similar pattern as the averaged trajectories in experiment 2. In order to be able to compare the two data sets, we will include all trajectories, including those where our simple algorithm did not detect a reset, as such trials are also included in the PSE data. Further, we proportionally scale the trajectories on both axes, so that the Y end-point matches the duration of stimulus. Since the PSE data was collected in degrees visual angle, we convert the unit to degrees visual angles, using the estimate at the centre of the workspace. Finally, for the 4 _s_ stimuli, we removed the last second. Figure 9 (top row) shows the PSE's obtained in the other study (shown separately for 2 and 4 DVA stimulus trajectories) as red dots, with the time-based model's average trajectories in black and the distribution of percepts as a gray histogram. The bottom row of Figure 9 shows the averaged trajectories across participants for each of the internal motions (purple, red, orange), and external motions (left and right), with the distribution of percepts as a gray histogram for the same 3 time points as in the top row.

For all internal motion speeds, we see a very similar average trajectory, that does not get closer to veridical over time. We verified this with a repeated measures ANOVA on perceived deviation from veridical stimulus position, using internal motion speed (2, 3, or 4 _cps_) and time (1, 2, or 3 _s_ after stimulus onset) as factors. If the trajectory data from experiment 2 here follows the same pattern as the PSE data, there should be an effect of time, and perhaps an interaction with internal motion speed. We find no effect of time ( _F_(2,16)=0.011, _p_=.957, $\eta^2$<.001), no effect of internal motion speed ( _F_(2,16)=0.127, _p_=.814, $\eta^2$=.0015) and no interaction ( _F_(4,32)=0.361, _p_=.635, $\eta^2$=.002). This means that we can't replicate the pattern found with the PSE approach. There may be several reasons for this. First, we notice that the trajectory data is highly dependent on participants. It could be that the re-tracing method provides very noisy estimates, hiding any effect. It could also be that the perception of the double-drift illusion, and it's resets, truly vary a lot between participants, so that both data sets and approaches are true. Since some participants do seem to follow the pattern found with the PSE approach, and some show a pattern of increasing average deviation between perceived and true position (not shown in the figure), it may that the differences between the data sets can simply be explained by random sampling from the population with relatively small numbers of participants.

@@

We first tested if tracking the position of a double-drift stimulus changes the illusory percept of it's position, and found that it doesn't. However, tracking the percept in real time did not show any unambiguous resets. To measure resets, e ask participants to draw the perceived path of the stimulus after a single pass. In this re-tracing experiment the drawn traces reflect the strength of the illusion as well, and in most trials there is a clear reset point, for either a jump reset or a hit-the-wall reset. We then set out to test if these spontaneous resets in the double-drift illusion reflect discrepancies between two position signals, occurring at some given size of this discrepancy, or if they simply occur after some time elapses, analogous to other illusions or bi-stable stimuli. The positions of these reset points do not vary much with elapsed time, but are better explained by the distance from the true position. This has implications for the way position perception relies on motion signals.

@@

Some of our participants seem to perceive "jump resets" though, and the model as published, does not predict any such resets. This could be added by including a causal inference step [e.g. @Wei2011]. For the subset of participants that experience jump resets, there may be a process that compares the two position signals (perceived and retinal) and when these exceed some threshold, decides that the two position signals do not share a common source. At that point the perceived position may be discarded, or reset to the retinal position, and the build-up of the illusion may start again. For the remaining participants, either the discrepancy between retinal and perceived location never exceeds their threshold, or their percept is indeed bounded to the maximum difference they allow between the two position signals.

Recently, it has been suggested that resets can be triggered by planned eye-movements to the stimulus [@Nakayama2020], particularly "jump" resets. In this experiment, we did not control for eye-movements with an eye-tracker, so we can't directly investigate this idea. Given a pre-motor theory of attention like mechanism, when spontaneously planned eye-movements trigger (jump) resets, such resets could occur at any point during the stimulus pass. However, it seems that for the participants that display "jump" resets in their re-tracing data, these happen at a fairly consistent location, which suggests that jump resets are instead usually triggered by the stimulus and not by attention. This should however be tested with eye-tracking data, so that this will remain a speculation here.
